Changes in Faculty Composition and Stagnating State Support for Higher Education

### 2023-03-19 text/sections/06-discussion.tex
Finished major revision.  Needs more checks of writing style.

The results show that the recent stagnation in state funding for higher education has affected the composition of faculty within public universities, away from tenure-track and tenured professors towards non-tenured lecturers.
At the same time, there are little discernible effects on individual professors hired in the years 2011-2021 at Illinois public universities, except for the salaries of first year lecturers.
The rate that faculty leave their university, and their rate of promotion between positions, are unaffected by the university revenues, so this leaves one primary channel to explain changes in faculty composition: hiring.

\cite{turner2014impact} documents the wide-spread practice of hiring freezes at universities in response to budget shocks around the 2008 recession.
Throughout the last decade, multiple such measures were taken by Illinois public universities in response to their deteriorating finances \citep{furlough2010}.
The University of Illinois\footnote{
    The University of Illinois includes three public university campuses: Urbana-Champaign, Chicago, Springfield.
}
did not receive the allocated state appropriations from the state of Illinois on time, so enacted cost-cutting measures to stay fiscally solvent.
The university system placed a hold on all hiring for filling state-funded positions and promotions.
Faculty were furloughed (placed on leave without pay) for a day each month, and university administrators placed on two days per month furlough, or eligible university employees could accept a voluntary, equivalent pay reduction.
The University of Western Illinois adopted very similar measures in response to the Illinois budget crisis in 2016-2017 \citep{wiu2016}.

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{Trends in New Hires at Illinois Public Universities 2011-2021.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{New Hire Count, Total.}
        \includegraphics[width=\textwidth]{figures/newhire-count-illinois.png}
        \label{fig:newhire-count-illinois}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Mean Salary in First Year for New Hires, \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/newhire-salary-illinois.png}
        \label{fig:newhire-salary-illinois}
    \end{subfigure}
    \label{fig:newhire-illinois}
\end{figure}

These examples show that Illinois public universities were effected by the negative shock to state funding for higher education, and instituted policy changes aimed at affecting faculty outcomes investigated in \autoref{sec:results}.
Yet the only measurable effect was on count of professor per student (implied via slower hiring), and first-year lecturer salaries.
To investigate these trends for the entire state, \autoref{fig:newhire-illinois} shows the count of professors hired in the years 2011-2021, and their mean salary, by position.
In particular, no position had sustained rises in hiring while enrolment within the university was rising (so that new hires per student necessarily fell), and hiring of new administrators jumped once the Illinois budget crisis ended in 2017.
At the same time, professor new hires' starting salaries remained relatively constant for lecturer and assistant professors, yet rose and fell over the decade for full professors and administrators.

>>>>

The results show that the recent stagnation in state funding for higher education has affected the composition of faculty within public universities, away from tenure-track and tenured professors towards contingent lecturers.
At the same time, there are little discernible effects on individual professors hired in the years 2011-2021 at Illinois public universities, except for the salaries of first year lecturers.
The rate that faculty leave their university, and their rate of promotion between positions, are unaffected by the university revenues, so this leaves one primary channel to explain changes in faculty composition: hiring.

Between academic years the number of professors at a university changes: professors leave an institution (retiring, hired elsewhere, etc.) and other join (they are newly hired, end a sabbatical, etc.).
\autoref{sec:results-indiv} shows that most of these channels are unaffected, implying that falls in hiring for tenured-track and tenured faculty must explain most of the substitution towards lecturers away from tenured faculty.
Notably, faculty hiring is already know to be heavily impacted by budget shocks.
\cite{turner2014impact} documents the wide-spread practice of hiring freezes at universities in response to budget shocks around the 2008 recession.
Throughout the last decade, multiple such measures were taken by Illinois public universities in response to their deteriorating finances \citep{furlough2010}.
The University of Illinois\footnote{
    The University of Illinois includes three public university campuses: Urbana-Champaign, Chicago, Springfield.
}
did not receive the allocated state appropriations from the state of Illinois on time, so enacted cost-cutting measures to stay fiscally solvent.
In response, the university system placed a hold on all hiring for filling state-funded positions and promotions.

The data used so far measure faculty count, so cannot disentangle which of these channels is most affected by changes in state funding at a public university, yet \cite{wapman2022quantifying} provide a measure of the aggregated total faculty hires at universities over the time period 2011--2021.\footnote{
    \citep{wapman2022quantifying} made an aggregated version of their data open-source, making this sub-analysis possible.
}
\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{State Funding and Faculty Hired at Public Universities, Total for 2011--2021.}
    \includegraphics[width=0.95\textwidth]{figures/hiring-correlation.png}
    \label{fig:hiring-correlation.png}
\end{figure}

Universities with more state funding (for the entire decade) hired more professors (\autoref{fig:hiring-correlation.png}).
Similarly, the funding shock instrument identifies that an increase in 10\% state funding leads to 13 more faculty hires across the decade 2011--2021, without any clear difference between the genders (\autoref{tab:hiring-shock-reg}).\footnote{
    These results were produced by integrating the total count of faculty hires for 2010--2021 for the top-ranked 180 US universities with a sum of the funding variables, and then estimating the models specified in \autoref{sec:iv-model-uni}.
    See \autoref{sec:appendix-hiring} for further details.
}
It is clear that public universities' hiring is negatively impacted by falls in state funding, even using these aggregated data for count of hiring. 
While this analysis strengthens the case negative effects on hiring plays, it is crude for measuring exact magnitudes for the effect on faculty hiring; I leave it to further research to delve deeper into the magnitudes that hiring plays as the primary causal mediator for the effects on faculty composition of falls in state funding for public universities.

### 2023-03-19 text/sections/07-conclusion.tex
Finished major revision.  Needs more checks of writing style.

This analysis investigates how the recent stagnation in state support for higher education has affected public universities, and their faculty composition.
The work contributes to the literature along two primary dimensions.
Firstly, by exploiting the same empirical approach as \cite{NBERw23736,NBERw27885} to isolate the causal effect of changes in state funding on public universities themselves, the analysis provides a plausible mechanism for the observed negative effects on students outcomes that this stagnation has brought.
Secondly, this analysis considers individual professors as the unit of analysis using a dataset (IBHED) new in the economic literature.
These data allow for detailed analysis of thousand's of professors' salaries, and are a starting point for further analysis of the determinants of professors wages in the Illinois public university system.

Findings show that public universities have substituted away from tenure-track and tenured professors, towards non-tenured lecturers, in the face of persistent declines in state funding for higher education.
The effects on incumbent professors in the state of Illinois are non-distinguishable from zero, which implies that the effect is driven by changes in university hiring of new professors.
Public universities are hiring thousands of non-tenured lecturers to satisfy teaching obligations, and fewer tenure-track and tenured professors that their private counterparts.

While costs education have been rising in the US, thanks to multiple overlapping trends \citep{ehrenberg2012}, public universities have also dealt with declining state support.
It is natural to expect that such headwinds will lead to systematic change at public universities, changes which are not neutral to the composition of faculty, or to their goals of research and education.
These results show large changes in faculty composition, and that stagnation in state support explains at least a third of the observed shift away from tenured professors and towards contingent lecturers.
At the same time, private universities were not exposed to financial headwinds of the same magnitude or persistence.
If public universities' hiring of faculty is disrupted by a fall in state funding, then we can expect that private universities will benefit (in a competitive sense) by being able to hire better professors from relatively resource constrained public universities.
While public universities continue to educate the majority of higher education students in the US, we should worry about the effects that restricting their financial support has on the composition of faculty, and the impact on higher education as a whole.

>>>>

This analysis investigates how the recent stagnation in state support for higher education has affected public universities, and their faculty composition.
This work contributes to the literature along two primary dimensions.
Firstly, by isolating changes in state funding on public universities via state funding shocks, the analysis provides a plausible mechanism for the observed negative effects on students outcomes that this stagnation has brought \citep{NBERw23736,NBERw27885}.
Secondly, this analysis considers individual professors as the unit of analysis using a dataset (IBHED) new in the economic literature.
These data allow for detailed analysis of thousand's of professors' salaries, and are a starting point for further analysis of the determinants of professors wages in the Illinois public university system.

Findings show that public universities have substituted away from tenure-track and tenured professors, towards non-tenured lecturers, in the face of persistent declines in state funding for higher education.
The effects on incumbent professors in the state of Illinois are non-distinguishable from zero, which implies that the effect is driven by changes in university hiring of new professors.
Each year, public universities are using more contingent lecturers to teach their education students, while private universities continue to employ more tenure-track and tenured professors than their public counterparts.
Private universities have always had more tenured professor per student than public institutions, but each year the gap widens.

While costs education have been rising in the US, thanks to multiple overlapping trends \citep{ehrenberg2012}, public universities have also dealt with declining state support.
It is natural to expect that such headwinds will lead to systematic change at public universities, changes which are not neutral to the composition of faculty, or to their goals of research and education.
These results show large changes in faculty composition, and that stagnation in state support explains at least a third of the observed shift away from tenured professors and towards contingent lecturers.
At the same time, private universities were not exposed to financial headwinds of the same magnitude or persistence.
If public universities' hiring of faculty is disrupted by a fall in state funding, then we can expect that private universities will benefit (in a competitive sense) by being able to hire better professors from relatively resource constrained public universities.
While public universities continue to educate the majority of higher education students in the US, we should worry about the effects that restricting their funding has on the composition of faculty, and the impact on higher education as a whole.

### 2023-03-19 text/sections/09-appendix.tex
Finished major revision.  Needs more checks of writing style.

@@

\begin{table}[h!]
    \singlespacing
    \centering
    \caption{IBHED Summary Statistics, Entire Professor Panel 2010--2021.}
    \makebox[\textwidth][c]{\input{tables/illinois-summary-rolling.tex}}
    \label{tab:illinois-summary-rolling}
    \begin{flushleft}
        \footnotesize
        \textbf{Note}: This table presents the summary statistics for professors with an observed year of first-year at their university.
    \end{flushleft}
\end{table}

@@

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{Local Projection Estimates for First-Stage \autoref{eqn:secondstage1_indiv}.}
    \includegraphics[width=0.6\textwidth]{figures/firststage-illinois-lp-rolling.png}
    \label{fig:firststage-illinois-lp}
\end{figure}

@@

>>>>

\subsection{First Stage Local Projection Estimates}

\begin{figure}[H]
    \centering
    \singlespacing
    \caption{Local Projection Estimates for First-Stage \autoref{eqn:firststage}.}
    \includegraphics[width=0.6\textwidth]{figures/firststage-lp.png}
    \label{fig:firststage-lp}
\end{figure}

@@

\begin{table}[H]
    \singlespacing
    \centering
    \caption{IBHED Summary Statistics, Entire Professor Panel 2010--2021.}
    \makebox[\textwidth][c]{\input{tables/illinois-summary-rolling.tex}}
    \label{tab:illinois-summary-rolling}
    \begin{flushleft}
        \footnotesize
        \textbf{Note}: This table presents the summary statistics for professors with an observed year of first-year at their university.
    \end{flushleft}
\end{table}

@@

\begin{figure}[H]
    \centering
    \singlespacing
    \caption{Local Projection Estimates for First-Stage \autoref{eqn:secondstage1_indiv}.}
    \includegraphics[width=0.6\textwidth]{figures/firststage-illinois-lp-rolling.png}
    \label{fig:firststage-illinois-lp}
\end{figure}

@@

\subsection{Additional Results, Faculty Hiring}
\label{sec:appendix-hiring}

These results were collects by using the open-source network data by \cite{wapman2022quantifying}, summing the count of hires a university made for 2011--2021 across the network structure of the data, and integrating a sum of all revenue related measures from IPEDS data.

\begin{table}[h!]
    \singlespacing
    \centering
    \caption{OLS and 2SLS Estimates for University Faculty Hiring, Total for 2011--2020.}
    \makebox[\textwidth][c]{\input{tables/hiring-shock-reg-fte.tex}}
    \begin{flushleft}
        \footnotesize
        \textbf{Note}: Standard errors are clustered at the state level.
    \end{flushleft}
    \label{tab:hiring-shock-reg}
\end{table}

Note that yearly variation is not observed here, so that only the aggregate level, for 180 universities, can be considered.
